Weak interspecific interactions in a sagebrush steppe: Evidence from observational data, population models, and a removal experiment

### 2016-09-30 removals/manuscript/removal_ms.tex
Minor removal ms edits

We simulated the IPM to estimate equilibrium abundances under three different scenarios. For all IPM simulations, we set the spatial random effects to zero, meaning that we are simulating dynamics on a hypothetical, average site. The first scenario featured all four species and our ``baseline'' model for control plots (no removal effects). We initialized all species with arbitrarily low cover and ran the model for 500 time steps (years) to allow the community to reach an equilibrium. To incorporate temporal environmental variation, at each time step we randomly selected one of the 30 sets of random year effects for the survival, growth and recruitment models (``kernel selection'' in the terminology of \citealt{metcalf_statistical_2015}). We calculated the equilibrium cover for each species by simulating an additional 2000 time steps and averaging each species' cover over this period. For the second scenario, we set the cover of \textit{A. tripartita} to zero and then repeated the procedure. This simulation estimates the competitive release of grasses following \textit{A. tripartita} removal according to the baseline model (no removal effects). For the third scenario, we again set  \textit{A. triparitia} to zero, but this time we projected the model using the vital rate regressions that include removal effects. If our baseline model successfully predicts competitive release observed in the removal experiment, then the results of the second and third scenarios should be similar. However, to the extent that the baseline model underestimates (or overestimates) the actual competitive release, the results of the second and third scenarios will differ.

As with the IBM, we repeated these simulations using maximum removal treatment effects represented by parameters values at the 5\% or 95\% confidence limit farthest from zero.

@@

@@

Our population models based on historical, observational data reveal weak interspecific interactions \citep{adler_coexistence_2010,chu_large_2015} and, as a result, predict very little competitive release among the four dominant species in a sagebrush steppe. Despite the fact that previous removal studies have reported variable results, ranging from negative effects of removal on remaining species to complete competitive release, we were suspicious of our model's predictions and conducted a removal experiment to test them. The results provide substantial, but not complete, support for our model's predictions.

@@

The small competitive release of many species in our system implies that each of the four dominant species we studied are limited by quite different factors. This low niche overlap could reflect strong resource partitioning (Tilman 1982), perhaps combined with variation in the times and locations at which species are taking up those resources. Species-specific responses to spatial and temporal variation in non-resource environmental variables can also weaken interspecific interactions \citep{chesson_mechanisms_2000}, though in a previous study we failed to find evidence of a temporal storage effect in this community \citep{adler_weak_2009}. Interactions with above- or belowground enemies can regulate populations and, given enough species-specificity in the effects of these enemies, decouple interspecific dynamics \citep{hersh_evaluating_2011,janzen_herbivores_1970,connell_role_1971}.  More likely, plants may alter the environment, and even resource availability, in ways that facilitate their neighbors. For example, shade cast by \textit{Artemisia} canopies may reduce evapotranspiration (e.g. \citealt{Barbier2008}). \textit{Artemisia} removal might offest this facilitative effect by reducing resource uptake, resulting in little net change in water balance following removal.  Unfortunately, we did not monitor soil moisture at the spatial and temporal resolution necessary to test this hypothesis.

@@

Our results have given us more confidence that competitive release is quite small for some species in some communities. But for many species in other communities, interspecific competition may still be an important driver. In closed-canopy forests where competition for light is intense, we might have found different results than in our system, in which competition for a number of belowground resources is more important than competition for light. However, a number of recent studies from forests have found that heterospecific competitive effects are dramatically weaker than conspecific effects \citep{comita_asymmetric_2010,kunstler_plant_2015,johnson_conspecific_2012}. We also expect stronger competitive release in successional systems, or communities in the early stages of community assembly where the process of competitive exclusion is still underway \citep{kokkoris_patterns_1999}. Finally, we might have found stronger competitive release if we could have studied rare species in our system, as long as those species are rare because they are competitively suppressed. If they are rare because they occupy unique, but narrow, niches, interspecific interactions would still be weak. Determining why interspecific interactions and competitive release are strong at some times and places and why they are weak in other situations remains a fascinating research problem.

@@

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% APPENDICES !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

>>>>

We simulated the IPM to estimate equilibrium abundances under three different scenarios. For all IPM simulations, we set the spatial random effects to zero, meaning that we are simulating dynamics on a hypothetical, average site. The first scenario featured all four species and our ``baseline'' model for control plots (no removal effects). We initialized all species with arbitrarily low cover and ran the model for 500 time steps (years) to allow the community to reach an equilibrium. To incorporate temporal environmental variation, at each time step we randomly selected one of the 30 sets of random year effects for the survival, growth and recruitment models (``kernel selection'' in the terminology of \citealt{metcalf_statistical_2015}). We calculated the equilibrium cover for each species by simulating an additional 2000 time steps and averaging each species' cover over this period. For the second scenario, we set the cover of \textit{A. tripartita} to zero and then repeated the procedure. This simulation estimates the competitive release of grasses following \textit{A. tripartita} removal according to the baseline model (no removal effects). For the third scenario, we again set  \textit{A. triparitia} to zero, but this time we projected the model using the vital rate regressions that include removal effects. If our baseline model successfully predicts competitive release observed in the removal experiment, then the results of the second and third scenarios should be similar. However, to the extent that the baseline model underestimates (or overestimates) the actual competitive release, the results of the second and third scenarios will differ. As with the IBM, we repeated these simulations using maximum removal treatment effects represented by parameters values at the 5\% or 95\% confidence limit farthest from zero. We did not simulate long-term \textit{A. tripartita} response to grass removal because the vital rate regressions for \textit{A. tripartita} showed that the baseline model effectively predicted removal effects.

@@

 The equilibrium cover of \text{Artemisia} simulated by the IPM (6.0\%) is consistent with observed cover during the historical data collection period (6.6\% from 1931-1956), but is low compared to contemporary observed cover (15-25\%,\ref{fig:CoverTrends}). Could the small competitive release following simulated \text{Artemisia} removal simply reflect the low \text{Artemisia} cover? To explore this possibility, we increased \text{Artemisia} cover in the baseline simulation by increasing average \text{Artemisia} fecundity. The increased fecundity led to a doubling of \text{Artemisia} average cover, but had essentially zero effect on the cover of the grasses (changes of less than 0.1\% cover).

@@

Our population models based on historical, observational data reveal weak interspecific interactions \citep{adler_coexistence_2010,chu_large_2015} and, as a result, predict very little competitive release among the four dominant species in a sagebrush steppe. Despite the fact that previous removal studies have reported variable results, ranging from negative effects of removal on remaining species to complete competitive release, we were suspicious of our model's predictions and conducted a removal experiment to test them. The results provide substantial support for our model's predictions.

@@

The small competitive release of many species in our system implies that each of the four dominant species we studied are limited by quite different factors. This low niche overlap could reflect strong resource partitioning (Tilman 1982), perhaps combined with variation in the times and locations at which species are taking up those resources. Species-specific responses to spatial and temporal variation in non-resource environmental variables can also weaken interspecific interactions \citep{chesson_mechanisms_2000}, though in a previous study we failed to find evidence of a temporal storage effect in this community \citep{adler_weak_2009}. Interactions with above- or belowground enemies can regulate populations and, given enough species-specificity in the effects of these enemies, decouple interspecific dynamics \citep{hersh_evaluating_2011,janzen_herbivores_1970,connell_role_1971}.  More likely, plants may alter the environment, and even resource availability, in ways that facilitate their neighbors. For example, shade cast by \textit{Artemisia} canopies may reduce evapotranspiration (e.g. \citealt{Barbier2008}). \textit{Artemisia} removal might offest this facilitative effect by reducing resource uptake, resulting in little net change in water balance following removal.  Unfortunately, we did not monitor soil moisture at the fine spatial and temporal resolution necessary to test this hypothesis.

@@

Our results have given us more confidence that competitive release is quite small for the dominant species in our study system. But for many species in other communities, interspecific competition may still be an important driver. In closed-canopy forests where competition for light is intense, we might have found different results than in our system, in which competition for a number of belowground resources is more important than competition for light. However, a number of recent studies from forests have also found that heterospecific competitive effects are dramatically weaker than conspecific effects \citep{comita_asymmetric_2010,kunstler_plant_2015,johnson_conspecific_2012}. We also expect stronger competitive release in successional systems, or communities in the early stages of community assembly where the process of competitive exclusion is still underway \citep{kokkoris_patterns_1999}. Finally, we might have found stronger competitive release if we could have studied rare species in our system, as long as those species are rare because of niche overlap with stronger competitors. If they are rare because they occupy unique, but narrow, niches, interspecific interactions would still be weak. Determining why interspecific interactions and competitive release are strong in some ecological contexts but weak in others remains a fascinating research problem.

@@

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% APPENDICES !
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
